Warming non-randomly speeds up food web fluxes in both an absolute and relative sense

### 2022-07-08 doc/ms/ms.Rmd
first update after prod fix. Push prior to taxa aggregatation in st7 & oh2

Warming temperatures are altering communities and trophic networks globally. While the influence of warming in food webs is often context-dependent, increasing temperatures are predicted to alter to alter food webs and ecosystems in a couple general and fundamental ways: 1) decreasing organism body size and 2) increasing metabolic rates. Both body size and metabolic rate are important drivers of many ecological patterns. Therefore, warming-induced changes have the potential to propagate through countless ecosystem processes thereby altering the distribution of food webs fluxes, food web stability, and potentially the importance of environmental and biological factors in community assembly. Here, we quantify the patterning and relative distribution of organic matter fluxes through stream food webs spanning a natural ~25$^\circ$C temperature gradient. We then related patterns in fluxes to species and community traits (body size, *M* and biomass turnover, *P:B*) within and across ecosystems. We predicted 1) fluxes would skew towards smaller *M* and higher *P:B* both across and within ecosystems and 2) fluxes would be more skewed at higher temperatures. Across the temperature gradient, communities in warmer streams were composed of smaller-bodied and higher-turnover populations on average. Additionally, material fluxes of warmer streams were increasingly skewed towards high-turnover populations *within* the community, showing higher temperatures restructured organic matter fluxes in both an absolute and relative sense. Lastly, in warmer ecosystems the relative distribution of organic matter fluxes appears to be increasingly 'non-random', evidence of potentially stronger selection for these traits with increasing temperature. The emerging picture is a warmer world that is both smaller and faster. Our study lends further support to this pattern and suggests that temperature will become an increasingly important environmental filter on warming communities in the future.
\newpage

@@

From modeled diet compositions, we estimated trophic redundancy within and across stream food webs by calculating proportional similarities (*PS*; @whittaker1952) among modeled diet estimates. Proportional similarities were calculated as:

@@

Consumption fluxes (g m^-2^ t^-1^) were calculated using the trophic basis of production method [TBP, @benke1980]. Taxon-specific secondary production estimates were combined with
diet proportions, diet-specific assimilation efficiencies, *AE~i~*, and assumed net production efficiencies, *NPE*, to estimate consumption of organic matter. For each food
category, *i*, diet proportions were multiplied by the gross growth efficiency ($GGE_{i} = AE_{i} * NPE$) to calculate the relative production attributable to each food
category. The relative production from each food type was then multiplied by the interval-level production and finally divided by $GGE_{i}$ to estimate consumption of organic
matter from each food category by a consumer [@benke1980]. Consumption was calculated for each taxon across sampling intervals (typically \~1 month). Total interval consumption was
calculated by summing across all taxa, while annual consumption was calculated by summing across all taxa and intervals. Variability in consumption estimates was estimated through
a Monte Carlo approach, wherein bootstrapped vectors of secondary production for each taxon (see *Secondary production* methods above) were resampled and consumption estimated with
the TBP method using modeled diet proportions (see *Diet analysis* above), diet-specific assimilation efficiencies, and net production efficiency. Variability in *AE~i~* was
incorporated by resampling values from beta distributions fit to median and 2.5% and 97.5% percentiles for each diet item: diatoms = 0.30 (95% percentile interval (PI): 0.24-0.36),
filamentous and green algae = 0.30 (95% PI: 0.24-0.36), cyanobacteria = 0.10 (95% PI: 0.08-0.12), amorphous detritus = 0.10 (95% PI: 0.08-0.12), vascular and non-vascular plants
(bryophytes) = 0.1 (95% PI:: 0.08-0.12), and animal material = 0.7 (95% PI: 0.56-0.84)[@welch1968; @cross2007; @cross2011; @benke1980; @benke1997]. Variability in *NPE* was
incorporated by resampling values from an assumed beta distribution with median *NPE* = 0.45 (95% PI = 0.4-0.5). Beta distributions were fit using the 'get.beta.par()' function
within the *rriskDistributions* package [@belgorodski2017].

@@

To assess the strength of environmental filtering of species' traits within a community, we quantified how likely it was to observe such a skewed distribution in OM fluxes by random chance. Here, we predicted that communities with strong filtering for traits would exhibit highly skewed OM distributions that would be unlikely due to chance (i.e., 'non-random ordering'). In contrast, weak environmental filtering would produce OM fluxes with skew values that are likely due to random chance ('random ordering'. To accomplish this, we first had to account for statistical constraints that restrict the range of possible outcomes [\ i.e., feasible set, @haegeman2008; @diaz2021a], given the number of species and the relative distribution of OM fluxes within a community. The number of unique orderings of species increases to computationally intractable numbers very quickly (e.g., $S!$, 10 species ~3.6e^6^ unique orderings). Therefore, we chose to permute a random subset of each stream community's feasible set by randomly ordering species and calculating the skewness in the cumulative distribution of annual OM fluxes 100,000 times in each stream. The number of random orderings was chosen as a balance between characterizing the distribution of skewness values within each feasible set and computational and time constrains. This permuted set allowed us to calculate the probability of observing the empirical skewness, $Sk_{flux}$, in each stream compared to a random ordering given the distribution of relative OM flux. To assess the relationship between the probability of 'non-random ordering' and temperature, we again used beta regression on estimates of the probability of seeing the empirical $Sk_{flux}$ value or one more extreme based on random ordering of species in the permuted feasible set. Here, values can take the range between 0 and 1 so no transformation was necessary. As necessary, when we observed many values near 0 and 1, we implemented a zero-one inflated beta regressions in brms [@burkner2017] to test if the probability of observing the empirical $Sk_{flux}$ was associated with mean annual temperature across streams. We predicted increasingly 'non-random' distributions in OM fluxes with increasing temperature.

@@

Across communities, average population body size, *M*, decreased and average biomass turnover, *P:B*, increased with increasing temperature (Figure 2). Mean population *M* decreased from `r ann_spp_summ %>% dplyr::select(M_mg_ind_mean) %>% max %>% round(2)` (`r ann_spp_summ %>% dplyr::slice_max(M_mg_ind_mean) %>% dplyr::select(M_mg_ind_quant2.5) %>% round(2)` -- `r ann_spp_summ %>% dplyr::slice_max(M_mg_ind_mean) %>% dplyr::select(M_mg_ind_quant97.5) %>% round(2)` 95% PI) in the coldest stream to `r ann_spp_summ %>% dplyr::select(M_mg_ind_mean) %>% min %>% round(2)` (`r ann_spp_summ %>% dplyr::slice_min(M_mg_ind_mean) %>% dplyr::select(M_mg_ind_quant2.5) %>% round(2)` -- `r ann_spp_summ %>% dplyr::slice_min(M_mg_ind_mean) %>% dplyr::select(M_mg_ind_quant97.5) %>% round(2)` 95% PI) in the warmest stream. This decrease corresponded to a `r ((exp(mean(m_spp_temp_coefs, na.rm = TRUE)) - 1)*100) %>% round(2)` (`r ((exp(quantile(m_spp_temp_coefs,0.025, na.rm = TRUE)) - 1)*100) %>% round(2)`--`r ((exp(quantile(m_spp_temp_coefs,0.975, na.rm = TRUE)) - 1)*100) %>% round(2)` 95% PI) percent decrease in mean population body size (mg ind^-1^) for every increase of 1$^\circ$C. In contrast, the average population *P:B* increased from `r ann_spp_summ %>% dplyr::select(pb_y_mean) %>% max %>% round(2)` (`r ann_spp_summ %>% dplyr::slice_max(pb_y_mean) %>% dplyr::select(pb_y_quant2.5) %>% round(2)` -- `r ann_spp_summ %>% dplyr::slice_max(pb_y_mean) %>% dplyr::select(pb_y_quant97.5) %>% round(2)` 95% PI) in the coldest stream to `r ann_spp_summ %>% dplyr::select(pb_y_mean) %>% min %>% round(2)` (`r ann_spp_summ %>% dplyr::slice_min(pb_y_mean) %>% dplyr::select(pb_y_quant2.5) %>% round(2)` -- `r ann_spp_summ %>% dplyr::slice_min(pb_y_mean) %>% dplyr::select(M_mg_ind_quant97.5) %>% round(2)` 95% PI) in the warmest stream. This increase corresponded to a `r ((exp(mean(pb_spp_temp_coefs, na.rm = TRUE)) - 1)*100) %>% round(2)` (`r ((exp(quantile(pb_spp_temp_coefs,0.025, na.rm = TRUE)) - 1)*100) %>% round(2)`--`r ((exp(quantile(pb_spp_temp_coefs,0.975, na.rm = TRUE)) - 1)*100) %>% round(2)` 95% PI) percent decrease in mean population body size (mg ind^-1^) for every increase of 1$^\circ$C (Figure 2).

@@

Organic matter fluxes within consumer communities were skewed towards larger body sizes (positive skew), towards smaller body sizes (negative skew), or neutrally in regards to body size among streams. Skew estimates with body size (*M*) ranging from `r m_skew_summ %>% slice_min(abs(M_mg_ind_skew_mean)) %>% select(M_mg_ind_skew_mean)%>% round(2)` to `r m_skew_summ %>% slice_max(abs(M_mg_ind_skew_mean)) %>% select(M_mg_ind_skew_mean) %>% round(2)` (Figure 3, Table 2). Similarly, skew in fluxes towards high *P:B* taxa varied among streams ranging from `r pb_skew_summ %>% slice_min(abs(pb_y_skew_mean)) %>% select(pb_y_skew_mean) %>% round(2)` to `r pb_skew_summ %>% slice_max(abs(pb_y_skew_mean)) %>% select(pb_y_skew_mean) %>% round(2)` (Figure 3, Table 2). The skew of OM fluxes towards populations with smaller/larger body sizes showed a small, yet systematic association with stream temperature (Figure 4a). The median effect size of increasing temperature was to shift the skewness coefficient, $Sk_{flux}$, `r M_skew_temp_coef_median` units (`r M_skew_temp_coef_quant2.5`--`r M_skew_temp_coef_quant97.5` 95% PI) for every 1$^\circ$C increase. Similarly, OM fluxes through the community skewed increasingly toward higher turnover (*P:B*) organisms with increasing temperature. The median increase in $Sk_{flux}$ towards higher *P:B* populations was `r pb_skew_temp_coef_mean` (`r pb_skew_temp_coef_quant2.5`--`r pb_skew_temp_coef_quant97.5` 95% PI; Figure 4c).

@@

@@

We are grateful to Sigurður Guðjonsson, Guðni Guðbergsson, and the staff at the Veiðimlastofnun for providing laboratory space and logistical support. We are also grateful to Sveinbj€orn Steinþorsson at the University of Iceland for super-jeep transport to our field sites during the winter. Jeff Wesner and Abe Kanz for code sharing and discussions on modeling diet proportions.

>>>>

Warming temperatures are altering communities and trophic networks globally. While the influence of warming in food webs is often context-dependent, increasing temperatures are predicted to alter food webs and ecosystems in a number of general and fundamental ways, among them, 1) decreasing organism body size and 2) increasing metabolic rates. Both body size and metabolic rate are important drivers of many ecological patterns. Therefore, warming-induced changes have the potential to propagate through countless ecosystem processes thereby altering the distribution of food webs fluxes, food web stability, and potentially the importance of environmental and biological factors in community assembly. Here, we quantify the patterning and relative distribution of organic matter fluxes through stream food webs spanning a natural ~25$^\circ$C temperature gradient. We then related patterns in fluxes to species and community traits (body size, *M* and biomass turnover, *P:B*) within and across ecosystems. We predicted 1) fluxes would skew towards smaller *M* and higher *P:B* both across and within ecosystems and 2) fluxes would be more skewed at higher temperatures. Across the temperature gradient, communities in warmer streams were composed of smaller-bodied and higher-turnover populations on average. Additionally, material fluxes of warmer streams were increasingly skewed towards high-turnover populations *within* the community, showing higher temperatures restructured organic matter fluxes in both an absolute and relative sense. Lastly, in warmer ecosystems the relative distribution of organic matter fluxes appears to be increasingly 'non-random', evidence of potentially stronger selection for these traits with increasing temperature. The emerging picture is a warmer world that is both smaller and faster. Our study lends further support to this pattern and suggests that temperature will become an increasingly important environmental filter on warming communities in the future.
\newpage

@@

From modeled diet compositions, we estimated trophic redundancy within and across stream food webs by calculating proportional similarities (*PS*; @whittaker1952 among modeled diet estimates. Proportional similarities were calculated as:

@@

Consumption fluxes (g m^-2^ t^-1^) were calculated using the trophic basis of production method [TBP, @benke1980]. Taxon-specific secondary production estimates were combined with
diet proportions, diet-specific assimilation efficiencies, *AE~i~*, and assumed net production efficiencies, *NPE*, to estimate consumption of organic matter. For each food
category, *i*, diet proportions were multiplied by the gross growth efficiency ($GGE_{i} = AE_{i} * NPE$) to calculate the relative production attributable to each food
category. The relative production from each food type was then multiplied by the interval-level production and finally divided by $GGE_{i}$ to estimate consumption of organic
matter from each food category by a consumer [@benke1980]. Consumption was calculated for each taxon across sampling intervals (typically \~1 month). Total interval consumption was
calculated by summing across all taxa, while annual consumption was calculated by summing across all taxa and intervals. Variability in consumption estimates was estimated through
a Monte Carlo approach, wherein bootstrapped vectors of secondary production for each taxon (see *Secondary production* methods above) were resampled and consumption estimated with
the TBP method using modeled diet proportions (see *Diet analysis* above), diet-specific assimilation efficiencies, and net production efficiency. Variability in *AE~i~* was
incorporated by resampling values from beta distributions fit to median and 2.5% and 97.5% percentiles for each diet item: diatoms = 0.30 (95% percentile interval (PI): 0.24-0.36),
filamentous and green algae = 0.30 (95% PI: 0.24-0.36), cyanobacteria = 0.10 (95% PI: 0.08-0.12), amorphous detritus = 0.10 (95% PI: 0.08-0.12), vascular and non-vascular plants
(bryophytes) = 0.1 (95% PI: 0.08-0.12), and animal material = 0.7 (95% PI: 0.56-0.84)[@welch1968; @cross2007; @cross2011; @benke1980; @benke1997]. Variability in *NPE* was
incorporated by resampling values from an assumed beta distribution with median *NPE* = 0.45 (95% PI = 0.4-0.5). Beta distributions were fit using the 'get.beta.par()' function
within the *rriskDistributions* package [@belgorodski2017].

@@

To assess the strength of environmental filtering of species' traits within a community, we quantified how likely it was to observe such a skewed distribution in OM fluxes by random chance. Here, we predicted that communities with strong filtering for traits would exhibit highly skewed OM distributions that would be unlikely due to chance (i.e., 'non-random ordering'). In contrast, weak environmental filtering would produce OM fluxes with skew values that are likely due to random chance ('random ordering'). To accomplish this, we first had to account for statistical constraints that restrict the range of possible outcomes [\ i.e., feasible set, @haegeman2008; @diaz2021a], given the number of species and the relative distribution of OM fluxes within a community. The number of unique orderings of species increases to computationally intractable numbers very quickly (e.g., $S!$, 10 species ~3.6e^6^ unique orderings). Therefore, we chose to permute a random subset of each stream community's feasible set by randomly ordering species and calculating the skewness in the cumulative distribution of annual OM fluxes 100,000 times in each stream. The number of random orderings was chosen as a balance between characterizing the distribution of skewness values within each feasible set and computational and time constrains. This permuted set allowed us to calculate the probability of observing the empirical skewness, $Sk_{flux}$, in each stream compared to a random ordering given the distribution of relative OM flux. To assess the relationship between the probability of 'non-random ordering' and temperature, we again used beta regression on estimates of the probability of seeing the empirical $Sk_{flux}$ value or one more extreme based on random ordering of species in the permuted feasible set. Here, values can take the range between 0 and 1 so no transformation was necessary. As necessary, when we observed many values near 0 and 1, we implemented a zero-one inflated beta regressions in brms [@burkner2017] to test if the probability of observing the empirical $Sk_{flux}$ was associated with mean annual temperature across streams. We predicted increasingly 'non-random' distributions in OM fluxes with increasing temperature.

@@

Across communities, average population body size, *M*, decreased and average biomass turnover, *P:B*, increased with increasing temperature (Figure 2). Mean population *M* decreased from `r ann_spp_summ %>% dplyr::select(M_mg_ind_mean) %>% max %>% round(2)` (`r ann_spp_summ %>% dplyr::slice_max(M_mg_ind_mean) %>% dplyr::select(M_mg_ind_quant2.5) %>% round(2)` -- `r ann_spp_summ %>% dplyr::slice_max(M_mg_ind_mean) %>% dplyr::select(M_mg_ind_quant97.5) %>% round(2)` 95% PI) in the coldest stream to `r ann_spp_summ %>% dplyr::select(M_mg_ind_mean) %>% min %>% round(2)` (`r ann_spp_summ %>% dplyr::slice_min(M_mg_ind_mean) %>% dplyr::select(M_mg_ind_quant2.5) %>% round(2)` -- `r ann_spp_summ %>% dplyr::slice_min(M_mg_ind_mean) %>% dplyr::select(M_mg_ind_quant97.5) %>% round(2)` 95% PI) in the warmest stream. This decrease corresponded to a `r ((exp(mean(m_spp_temp_coefs, na.rm = TRUE)) - 1)*100) %>% round(2)` (`r ((exp(quantile(m_spp_temp_coefs,0.025, na.rm = TRUE)) - 1)*100) %>% round(2)`--`r ((exp(quantile(m_spp_temp_coefs,0.975, na.rm = TRUE)) - 1)*100) %>% round(2)` 95% PI) percent decrease in mean population body size (mg ind^-1^) for every increase of 1$^\circ$C. In contrast, the average population *P:B* increased from `r ann_spp_summ %>% dplyr::select(pb_y_mean) %>% min %>% round(2)` (`r ann_spp_summ %>% dplyr::slice_min(pb_y_mean) %>% dplyr::select(pb_y_quant2.5) %>% round(2)` -- `r ann_spp_summ %>% dplyr::slice_min(pb_y_mean) %>% dplyr::select(pb_y_quant97.5) %>% round(2)` 95% PI) in the coldest stream to `r ann_spp_summ %>% dplyr::select(pb_y_mean) %>% max %>% round(2)` (`r ann_spp_summ %>% dplyr::slice_max(pb_y_mean) %>% dplyr::select(pb_y_quant2.5) %>% round(2)` -- `r ann_spp_summ %>% dplyr::slice_max(pb_y_mean) %>% dplyr::select(M_mg_ind_quant97.5) %>% round(2)` 95% PI) in the warmest stream. This increase corresponded to a `r ((exp(mean(pb_spp_temp_coefs, na.rm = TRUE)) - 1)*100) %>% round(2)`% (`r ((exp(quantile(pb_spp_temp_coefs,0.025, na.rm = TRUE)) - 1)*100) %>% round(2)`--`r ((exp(quantile(pb_spp_temp_coefs,0.975, na.rm = TRUE)) - 1)*100) %>% round(2)` 95% PI) increase in mean population biomass turnover (P:B y^-1^) for every increase of 1$^\circ$C (Figure 2).

@@

Organic matter fluxes within consumer communities were skewed towards larger body sizes (positive skew), towards smaller body sizes (negative skew), or neutrally in regards to body size among streams. Skew estimates with body size (*M*) ranging from `r m_skew_summ %>% slice_min(M_mg_ind_skew_mean) %>% select(M_mg_ind_skew_mean)%>% round(2)` to `r m_skew_summ %>% slice_max(M_mg_ind_skew_mean) %>% select(M_mg_ind_skew_mean) %>% round(2)` (Figure 3, Table 2). Similarly, skew in fluxes towards high *P:B* taxa varied among streams ranging from `r pb_skew_summ %>% slice_min(pb_y_skew_mean) %>% select(pb_y_skew_mean) %>% round(2)` to `r pb_skew_summ %>% slice_max(pb_y_skew_mean) %>% select(pb_y_skew_mean) %>% round(2)` (Figure 3, Table 2). The skew of OM fluxes towards populations with smaller/larger body sizes showed no systematic association with stream temperature (Figure 4a). Similarly, OM fluxes through the community skewed increasingly toward higher turnover (*P:B*) organisms with increasing temperature. The median increase in $Sk_{flux}$ towards higher *P:B* populations was `r pb_skew_temp_coef_mean` (`r pb_skew_temp_coef_quant2.5`--`r pb_skew_temp_coef_quant97.5` 95% PI; Figure 4c).

@@

Temperature has long been known to be an important control on ecological process rates [@arrhenius1897; @gillooly2003; @brown2004]

@@

We are grateful to Sigurður Guðjonsson, Guðni Guðbergsson, and the staff at the Veiðimlastofnun for providing laboratory space and logistical support. We are also grateful to Sveinbj€orn Steinþorsson at the University of Iceland for super-jeep transport to our field sites during the winter. We thank Lauren David, David Hernandez, Amanda Keasberry, Elena Nava, Camille Perrett, Jackie Pitts, Friðþjófur Árnason, Liliana García, Ragnahildur Magnúsdottír, Ryan McClure, Vija Pelekis, Adam Toomey, Chau Tran, Brooke Weigel, Tanner Williamson and many undergraduate workers for field and laboratory help. Jeff Wesner and Abe Kanz generously provided R code and discussions on modeling diet proportions. This study was supported by the National Science Foundation (DEB-0949774 and DEB-1354624 to JPB and ADH and DEB-0949726 to WFC).
